RationalGRL: A Framework for Argumentation and Goal Modeling

### 2018-09-17 discussion.tex
Section 6-7-8

\subsection{Related work}
\label{sect:discussions:relatedwork}

@@

Another recent example of the use of arguments in goal-oriented requirements engineering is the work by Murukannaiah \emph{et al.}~\cite{murukannaiah2015}, who propose Arg-ACH, where the beliefs underlying conflicting goals can be made explicit using argumentation. Murukannaiah \emph{et al.} start with the basic technique of Analysis of Competing Hypotheses (ACH), where for conflicting goals the beliefs that are consistent and inconsistent with these goals are included in a matrix and counted. They then extend this technique into Arg-ACH: instead of just indicating whether a belief is consistent or inconsistent with a goal, each belief becomes an argument for or against the goal, which is then diagrammed using the Carneades tool \cite{gordon2007visualizing}. Belief scores are assigned to arguments, which can be aggregated to provide one's belief in a goal. The arguments for and against goals can be based on argument schemes, and critical questions can be used to find new arguments for or against the goals or the existing arguments. One example provided in the paper is the argument scheme from expert opinion, which allows one to draw conclusions based on expert statements and subsequently question, for example, the objectivity and veracity of the expert using critical questions. Murukannaiah \emph{et al.} conducted an experiment in which they had two groups, one with ACH and one with Arg-ACH, perform an analysis of several conflicting goals regarding security at transport hubs. They found that, while the group that used Arg-ACH took longer, they also covered more possibilities in their belief search and the conclusions were more consistent among the group.

@@

More explicit rationales for goal models are included in the original GRL specification \cite{Amyot:2010:EGM:1841349.1841356} in the form of \emph{belief} elements, which can be attached to other intentional elements (goals, tasks) using the different links (positive contribution, negative contribution) of GRL. The content of a belief is a textual description, which can explain, for instance, the rationale for including a goal in a model. In this regard, they seem like arguments or reasons for the different elements in a goal model. However, in the literature the precise usage and semantics of belief elements in GRL remain unclear: according to the metamodel \cite{Amyot:2010:EGM:1841349.1841356} they can be connected to other intentional elements (including beliefs) through any type of link, but in the jUCMNav tool there is a separate \emph{belief link} for connecting beliefs to goals and tasks only, and the beliefs are not used in any of the algorithms to compute the satisfiability of goals and tasks. We are also not aware of many examples of the usage of belief elements in the literature. So what it seems like is that GRL belief elements can be understood as metadata that can be attached to an element

Ambiguities notwithstanding, there are further advantages of using RationalGRL over GRL with belief elements. For example, in none of the GRL specifications is it possible to attach beliefs to links between other elements so it is not possible, for example, to provide a rationale for why some tasks positively contributes to a goal. Furthermore, the RationalGRL arguments follow clear schemes with associated critical questions explicitly tailored towards reasoning about goals and actions or tasks, while beliefs in GRL offer no further guidance as to what kinds of rationales can be given for elements in a goal model. Furthermore, beliefs in GRL seem not to influence the goal model in any way. Arguments in RationalGRL, in contrast, play an important role in forming the resulting goal model: if an element is defeated by an argument, it is not part of the resulting goal model.

@@

There are several contributions to the literature that explicitly relate argumentation-based techniques with goal modeling. The first line of work is by Bagheri and Ensan \cite{bagheri2011consolidating} and Mirbel and Villata \cite{MirbelVillata12} use abstract argumentation frameworks (see Definition~\ref{def:argumentation-framework}) to capture individual goals and the relations between them as modeled in a goal model, so that consistent (i.e., acceptable) subsets of goals can be computed using the relevant argumentation semantics \cite{Dung1995}. For example, if in the goal model there is a conflict between goals $G_1$ and $G_2$, there is an attack between the arguments representing these goals and hence $G_1$ and $G_2$ cannot be in the same extension. Or if there is a dependency link between $G_1$ and $G_2$, then $G_1$ should be in every extension $G_2$ is in and vice versa. AND and OR decomposition are also modeled as such, that is, if $G_3$ AND-decomposes into $G_1$ and $G_2$ then $G_1$ and $G_2$ should be in every extension $G_3$ is in, and if $G_3$ OR-decomposes into $G_1$ and $G_2$ then either $G_1$ or $G_2$ should be in every extension $G_3$ is in.

@@

Additionally, from the user evaluation of the tool it followed that in order to implement the tool in realistic settings, other ways may have to be found to incorporate arguments in the goal modeling process. One idea is to use the tool for collaborative on-line goal modeling, similar to Github: since the renaming (i.e. replacement) and deletion (i.e. disabling) of elements are all logged, it is easy for a stakeholder to continue working on a model that was made by another stakeholder. Furthermore, critical questions can be included for other stakeholders to answer at a later date. As a formal underpinning of this asynchronous communication between users, it would make sense to capture requirements engineering and software design processes as dialogs between parties~\cite{finkelstein1989multiparty,BlackEtal2013}, which are a natural fit with the question-answer format employed in the RationalGRL framework. Finally, it would make sense to further develop the RationalGRL Development Process. One idea here is to have stakeholders play a card game (similar to the well-known planning poker) with cards that contain critical, reflective questions (cf. \cite{TangEtal2018}).

>>>>

\subsection{Related Work}
\label{sect:discussions:relatedwork}

@@

Another recent example of the use of arguments in goal-oriented requirements engineering is the work by Murukannaiah \emph{et al.}~\cite{murukannaiah2015}, who propose Arg-ACH, where the beliefs underlying conflicting goals can be made explicit using argumentation. Murukannaiah \emph{et al.} start with the basic technique of Analysis of Competing Hypotheses (ACH), where for conflicting goals, the beliefs that are consistent and inconsistent with these goals are included in a matrix and counted. They then extend this technique into Arg-ACH: instead of just indicating whether a belief is consistent or inconsistent with a goal, each belief becomes an argument for or against the goal, which is then diagrammed using the Carneades tool \cite{gordon2007visualizing}. Belief scores are assigned to arguments, which can be aggregated to provide one's belief in a goal. The arguments for and against goals can be based on argument schemes, and critical questions can be used to find new arguments for or against the goals or the existing arguments. One example provided in the paper is the argument scheme from expert opinion, which allows one to draw conclusions based on expert statements and subsequently question, for example, the objectivity and veracity of the expert using critical questions. Murukannaiah \emph{et al.} conducted an experiment in which they had two groups, one with ACH and one with Arg-ACH, perform an analysis of several conflicting goals regarding security at transport hubs. They found that, while the group that used Arg-ACH took longer, they also covered more possibilities in their belief search and the conclusions were more consistent among the group.

@@

More explicit rationales for goal models are included in the original GRL specification \cite{Amyot:2010:EGM:1841349.1841356} in the form of \emph{belief} elements, which can be attached to other intentional elements (goals, tasks) using the different links (positive contribution, negative contribution) of GRL. The content of a belief is a textual description, which can explain, for instance, the rationale for including a goal in a model. In this regard, they seem like arguments or reasons for the different elements in a goal model. However, in the literature the precise usage and semantics of belief elements in GRL remain unclear: according to the metamodel \cite{Amyot:2010:EGM:1841349.1841356} they can be connected to other intentional elements (including beliefs) through any type of link, but in the jUCMNav tool there is a separate \emph{belief link} for connecting beliefs to goals and tasks only, and the beliefs are not used in any of the algorithms to compute the satisfiability of goals and tasks. We are also not aware of many examples of the usage of belief elements in the literature. Thus, what it seems like is that GRL belief elements can be understood as metadata that can be attached to an element.

Ambiguities notwithstanding, there are further advantages of using RationalGRL over GRL with belief elements. For example, in none of the GRL specifications, is it possible to attach beliefs to links between other elements thus it is not possible, for example, to provide a rationale for why some tasks positively contributes to a goal. Furthermore, the RationalGRL arguments follow clear schemes with associated critical questions explicitly tailored towards reasoning about goals and actions or tasks, while beliefs in GRL offer no further guidance as to what kinds of rationales can be given for elements in a goal model. Furthermore, beliefs in GRL seem not to influence the goal model in any way. Arguments in RationalGRL, in contrast, play an important role in forming the resulting goal model: if an element is defeated by an argument, it is not part of the resulting goal model.

@@

There are several contributions to the literature that explicitly relate argumentation-based techniques with goal modeling. The first line of work is by Bagheri and Ensan \cite{bagheri2011consolidating} and Mirbel and Villata \cite{MirbelVillata12} use abstract argumentation frameworks (see Definition~\ref{def:argumentation-framework}) to capture individual goals and the relations between them as modeled in a goal model, so that consistent (i.e., acceptable) subsets of goals can be computed using the relevant argumentation semantics \cite{Dung1995}. For example, if in the goal model there is a conflict between goals $G_1$ and $G_2$, there is an attack between the arguments representing these goals and hence $G_1$ and $G_2$ cannot be in the same extension. Or if there is a dependency link between $G_1$ and $G_2$, then $G_1$ should be in every extension $G_2$ is in and vice versa. AND- and OR-decomposition are also modeled as such, that is, if $G_3$ AND-decomposes into $G_1$ and $G_2$ then $G_1$ and $G_2$ should be in every extension $G_3$ is in, and if $G_3$ OR-decomposes into $G_1$ and $G_2$ then either $G_1$ or $G_2$ should be in every extension $G_3$ is in.

@@

Additionally, from the user evaluation of the tool, it followed that in order to implement the tool in realistic settings, other ways may have to be found to incorporate arguments in the goal modeling process. One idea is to use the tool for collaborative on-line goal modeling, similar to Github: since the renaming (i.e. replacement) and deletion (i.e. disabling) of elements are all logged, it is easy for a stakeholder to continue working on a model that was made by another stakeholder. Furthermore, critical questions can be included for other stakeholders to answer at a later date. As a formal underpinning of this asynchronous communication between users, it would make sense to capture requirements engineering and software design processes as dialogs between parties~\cite{finkelstein1989multiparty,BlackEtal2013}, which are a natural fit with the question-answer format employed in the RationalGRL framework. Finally, it would make sense to further develop the RationalGRL Development Process. One idea here is to have stakeholders play a card game (similar to the well-known planning poker) with cards that contain critical, reflective questions (cf. \cite{TangEtal2018}).

### 2018-09-17 validation.tex
Section 6-7-8

\section{RationalGRL evaluation experiment}
\label{sect:validation}

In addition to the empirical case study (Section~\ref{sect:gmas}), which forms the basis of the RationalGRL framework, we also performed a user evaluation with 16 users. The objective of this evaluation was to determine whether the users found the additions of RationalGRL (arguments, argumentation semantics, critical questions) useful compared to standard GRL, and whether they found it easier to keep track of and express opinions and beliefs using RationalGRL compared to standard GRL. Regarding the comparison of RationalGRL to GRL, the following two broad hypotheses follow from our main points in this paper:

\begin{itemize}
\item[H1] The additions of RationalGRL -- arguments, critical questions and determination of the status of arguments -- are a useful addition to standard goal modelling in GRL.
\item[H2] The additions of RationalGRL make it easier to express, determine the effect of and communicate to other stakeholders one's opinions and beliefs about a goal model.
\end{itemize}

In addition, we wanted to know whether the RationalGRL tool was difficult or easy to use. Note that we did not aim to conduct a usability study for the RationalGRL tool. Rather, we were interested to know if the tool was clearly very easy or difficult to use, as this could influence the participants ideas about the RationalGRL framework and language -- a very nice and easy to use tool tends to make people more positive towards a particular modelling language, whereas irritations about a bad tool will lead to a more negative disposition. This leads to the following general hypothesis:

@@

The idea was to have our participants perform a small modelling task, and then ask them what they thought of RationalGRL and the tool. Thus, our experiment consisted of three parts:

@@

The instructions for these three parts are contained in a single document that we sent out to the participants.\footnote{See \url{www.rationalgrl.com}, page ``Empirical Study''} The explanation (part 1) starts with a general explanation of GRL and standard goal modelling similar to the explanation provided in section \ref{sect:background:grl}, then briefly discusses the additions of RationalGRL (argument, attack, critical questions) and finally the tool is briefly described (similar to Section \ref{sect:tool}).

@@

The second part of the survey concerns the comparison between GRL and RationalGRL and starts with Figure~\ref{fig:example-small}, which models the discussion for part 2 (Appendix~\ref{sect:survey}) in standard GRL, so that the participants can compare their own RationalGRL models and experiences with the standard GRL model. The following questions ask the participants to rate RationalGRL on a Likert scale from 1 (very useless) to 5 (very useful):
\begin{itemize}
\item[Q5] Do you think the arguments and counterarguments of RationalGRL are a useful extension to standard goal modeling?
\item[Q6] Do you think the critical questions and answers in the details pane are a useful extension to standard goal modeling?
\item[Q7] Do you think the automatic determination of the status of arguments and elements is a useful extension to standard goal modeling?
\end{itemize}
The following questions ask the participants to rate RationalGRL vs. standard goal modelling on a Likert scale from 1 (much more difficult) to 5 (much easier):
\begin{itemize}
\item[Q8] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone to express beliefs and opinions in a goal model?
\item[Q9] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone to determine the effect of beliefs and opinions on the resulting goal model?
\item [Q10] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone who is not the original author to understand the goal model? 
\end{itemize}
We conclude with two open questions regarding the strengths and weaknesses of RationalGRL compared to GRL.

In order to test our main hypotheses H1, H2 and H3, we have to formulate the appropriate null hypotheses and alternative hypotheses. For H1 and H2, we hypothesize that the participants will, on average, rate the additions of RationalGRL as useful (rating 4) or very useful (rating 5) for Q5, Q6 and Q7, and as making it easier (rating 4) or significantly easier (rating 5) for questions Q8, Q9 and Q10. In other words, we say that our null hypothesis and alternative hypothesis for H1 and H2 are as follows:
\begin{itemize}
\item H1$_{0}$ and H2$_{0}$: average rating for questions Q5-Q10 is 3 or lower.
\item H1$_{a}$ and H2$_{a}$: average rating for questions Q5-Q10 is higher than 3.
\end{itemize}
Note that for each question (Q5-Q10) we can test the null and alternative hypothesis separately, giving insight into exactly which additions of RationalGRL were deemed useful.

@@

\paragraph{Participants}
We asked 16 participants in our network to participate in our experiment. Most of the participants were therefore either employed at the first author's company or staff and (ex-)students from the second author's university. Hence, two thirds of the participants had either a PhD or Master degree, 20\% a Bachelor degree, and one respondent responded with ``Other''. All but one participant had a year or more experience with software development. The average experience was 6.2 years (standard deviation 6.7), with 10 participants having less than 5 years of experience and 6 participants having 8 or more years of experience, with one participant having 25 years of experience. The participants also judged themselves to be quite competent in  early-phase requirements engineering: on average they gave themselves a rating of 3.2 out of 5 (standard deviation 1.3) with half of the participants claiming they were at least ``competent'' (rating 4) or ``very competent'' (rating 5). However, experience with goal modeling languages was markedly less: the average rating was 1.9 (standard deviation 1.3), with 9 participants claiming never to have used a goal modeling language (rating 1), and only two participants displaying regular use (monthly, rating 4, or weekly, rating 5). The participants that had experience with goal modelling languages had mostly used i* (5 users), with 2 users being familiar with GRL and 2 users having used another goal modelling language.

@@

After the modelling task, the survey questions were asked. Table~\ref{table:survey:table1} shows the respondents' answers to question Q1-Q4 together with some statistical results which will be further discussed in Section \ref{sect:validation:analysis}.

@@

\subsection{Analysis or results}\label{sect:validation:analysis}

\paragraph{Hypothesis H1}
The first analysis concerns hypothesis H1, whether the additions of RationalGRL -- arguments, critical questions and determination of the status of arguments -- are a useful addition to standard goal modelling in GRL. Table~\ref{table:survey:table2} summarizes the average scores, standard deviations, p-values for a right-tailed t-test, effect size and statistical power for the relevant questions Q5-Q7. As indicated by the p-values, which are all lower than 0.05, all the averages are significantly higher than the median of 3 with a significance level ($\alpha$) of 0.05, which means that we can accept H1$_{a}$ and reject H1$_{0}$ for all the separate questions Q5-Q7. In other words, the fact that the participants rate the additions of RationalGRL as, on average, (very) useful is very unlikely to be due to chance.

@@

\paragraph{Hypothesis H3}
The third hypothesis H3 is about the usability of the RationalGRL tool. It is different from the other two hypotheses in that we do not expect or claim that the usability of the tool is very good (or bad). Rather, because in the RationalGRL framework the tool is more a means than an end in itself, we hypothesize that the too, is neither easy or difficult to use. Table~\ref{table:survey:table1} summarizes the average scores, standard deviations and p-values for a two-tailed t-test for questions Q1-Q4. As indicated by the p-values which are all higher than $\alpha=0.05$, none of the results is statistically significant, that is, the fact that there is a difference between the median of 3 and the average scores for the questions could very well be due to chance. This means we cannot reject the null hypothesis H3$_{0}$ and that, hence, the usability of the RationalGRL tool is neither (very) easy or (very) difficult. This corresponds to the answers to the open questions the participants gave: respondents were positive about the fact that it was very easy to get started with the tool and most comments on the tool were the type of comments that are expected from a prototype implementation.

\subsection{Threats to validity}
Internal validity is about the validity of the experiment results given our experiment setup and interpretation of findings. The main threat in this regard is that the modeling task the participants were asked to do is not realistic. The task is quite small, and based on an existing transcript of a requirements discussion. In a realistic setting, users of RationalGRL would model much more complex requirements together with the stakeholders. That said, the discussions analysed in our case study (Section \ref{sect:gmas}) were of a realistic size and included many arguments and critical questions, so these concepts seem to take a central role in goal modeling and requirements engineering.

Another threat to internal validity is that many of the participants had no experience with goal modelling languages at all. Hence, the perceived usefulness of the arguments could be influenced by the fact that the participants, for the first time, were given a goal modeling language to perform structured requirements analysis. In order to test the usefulness of arguments and critical questions alone, one would have to compare a test and control group, where the test group uses RationalGRL and the control group uses normal GRL.

@@

\begin{enumerate}
\item \emph{Arguments and automatically determining acceptability are useful.} Participants were enthusiastic about adding arguments to a goal modeling language with a clear formal underpinning, and they believed the argumentation semantics we use in the tool are intuitive. This is a positive signal for the addition of arguments to goal modeling languages, and also for our formal approach, as just adding arguments as, for example, labels on goals and tasks does not allow one to compute the status of elements in a goal model in the way our approach does. Furthermore, automatically determining the status of elements will perhaps have a bigger impact when working with larger, more complex goal models, for which manually determining the impact of arguments is not feasible.
\item \emph{Critical questions need to be clear and have a prominent place in the process of building a goal model.} While participants overall found the critical questions (CQ's) useful, their impact was markedly less than the addition of arguments. This can have various causes. Participants mentioned they did not use the details panel of the tool, which can be used to ask and answer critical questions, that much. Furthermore, it was also mentioned that the difference between different questions was not always clear. Finally, the experiment task, modeling an existing discussion, may also have influenced the CQ's perceived usefulness. The CQ's are meant to be asked during the actual RationalGRL Development Process in which goal models are constructed and critically analysed from scratch. 
\item \emph{High cognitive overhead.} A concern that was raised often in our empirical evaluation is that, in its current form, RationalGRL has a relatively high cognitive overhead. Goal modeling is by itself already a cognitively high-effort activity, and the fact that we add more elements to the language does not improve this. 
\end{enumerate}

>>>>

\section{RationalGRL Evaluation Experiment}
\label{sect:validation}

In addition to the empirical case study (Section~\ref{sect:gmas}), which forms the basis of the RationalGRL framework, we also performed a user evaluation with 16 users. The objective of this evaluation was to determine whether the users found the additions of RationalGRL (arguments, argumentation semantics, critical questions) useful compared to the standard GRL, and whether they found it easier to keep track of and express opinions and beliefs using RationalGRL compared to the standard GRL. Regarding the comparison of RationalGRL to GRL, the following two broad hypotheses follow from our main points in this paper:

\begin{itemize}
\item[H1] The additions of RationalGRL -- arguments, critical questions and determination of the status of arguments -- are a useful addition to standard goal modeling in GRL.
\item[H2] The additions of RationalGRL make it easier to express, determine the effect of and communicate to other stakeholders one's opinions and beliefs about a goal model.
\end{itemize}

In addition, we wanted to know whether the RationalGRL tool was difficult or easy to use. Note that we did not aim to conduct a usability study for the RationalGRL tool. Rather, we were interested to know if the tool was clearly very easy or difficult to use, as this could influence the participants ideas about the RationalGRL framework and language -- a very nice and easy to use tool tends to make people more positive towards a particular modeling language, whereas irritations about a bad tool will lead to a more negative disposition. This leads to the following general hypothesis:

@@

The idea was to have our participants perform a small modeling task, and then ask them what they thought of RationalGRL and the tool. Thus, our experiment consisted of three parts:

@@

The instructions for these three parts are provided in a single document that we sent out to the participants.\footnote{See \url{www.rationalgrl.com}, page ``Empirical Study''} The explanation (part 1) starts with a general explanation of GRL and standard goal modeling similar to the explanation provided in Section \ref{sect:background:grl}, then briefly discusses the additions of RationalGRL (argument, attack, critical questions) and finally the tool is briefly described (similar to Section \ref{sect:tool}).

@@

The second part of the survey concerns the comparison between GRL and RationalGRL and starts with Figure~\ref{fig:example-small}, which models the discussion for part 2 (Appendix~\ref{sect:survey}) in standard GRL, so that the participants can compare their own RationalGRL models and experiences with the standard GRL model. The following questions ask the participants to rate RationalGRL on a Likert scale from 1 (very useless) to 5 (very useful):
\begin{itemize}
\item[Q5] Do you think the arguments and counterarguments of RationalGRL are a useful extension to standard goal modeling?
\item[Q6] Do you think the critical questions and answers in the details pane are a useful extension to standard goal modeling?
\item[Q7] Do you think the automatic determination of the status of arguments and elements is a useful extension to standard goal modeling?
\end{itemize}
The following questions ask the participants to rate RationalGRL vs. standard goal modeling language  on a Likert scale from 1 (much more difficult) to 5 (much easier):
\begin{itemize}
\item[Q8] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone to express beliefs and opinions in a goal model?
\item[Q9] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone to determine the effect of beliefs and opinions on the resulting goal model?
\item [Q10] Do you think using RationalGRL instead of a standard goal modeling language makes it easier or more difficult to for someone who is not the original author to understand the goal model? 
\end{itemize}
We conclude with two open questions regarding the strengths and weaknesses of RationalGRL compared to GRL.

In order to test our main hypotheses H1, H2 and H3, we have to formulate the appropriate null hypotheses and alternative hypotheses. For H1 and H2, we hypothesize that the participants will, on average, rate the additions of RationalGRL as useful (rating 4) or very useful (rating 5) for Q5, Q6 and Q7, and as making it easier (rating 4) or significantly easier (rating 5) for questions Q8, Q9 and Q10. In other words, we say that our null hypothesis and alternative hypothesis for H1 and H2 are as follows:
\begin{itemize}
\item H1$_{0}$ and H2$_{0}$: average rating for questions Q5-Q10 is 3 or lower.
\item H1$_{a}$ and H2$_{a}$: average rating for questions Q5-Q10 is higher than 3.
\end{itemize}
Note that for each question (Q5-Q10,) we can test the null and alternative hypothesis separately, giving insight into exactly which additions of RationalGRL were deemed useful.

@@

\paragraph{Participants}
We asked 16 participants in our network to participate in our experiment. Most of the participants were therefore either employed at the first author's company or staff and (ex-)students from the second author's university. Hence, two thirds of the participants had either a PhD or Master degree, 20\% a Bachelor degree, and one respondent responded with ``Other''. All but one participant had a year or more experience with software development. The average experience was 6.2 years (standard deviation 6.7), with 10 participants having less than 5 years of experience and 6 participants having 8 or more years of experience, with one participant having 25 years of experience. The participants also judged themselves to be quite competent in  early-phase requirements engineering: on average they gave themselves a rating of 3.2 out of 5 (standard deviation 1.3) with half of the participants claiming they were at least ``competent'' (rating 4) or ``very competent'' (rating 5). However, experience with goal modeling languages was markedly less: the average rating was 1.9 (standard deviation 1.3), with 9 participants claiming never to have used a goal modeling language (rating 1), and only two participants displaying regular use (monthly, rating 4, or weekly, rating 5). The participants that had experience with goal modeling languages had mostly used i* (5 users), with 2 users being familiar with GRL and 2 users having used another goal modeling language.

Note that the third author of the paper did not participate in recruiting of the participants, data collection and data processing of the user study due to the limitation of ``Institutional Review Board for the Protection of Human Subjects (IRB)" at her home university. Thus, she has not seen the participants information, the raw and processed data.

@@

After the modeling task, the survey questions were asked. Table~\ref{table:survey:table1} shows the respondents' answers to question Q1-Q4 together with some statistical results which will be further discussed in Section \ref{sect:validation:analysis}.

@@

\subsection{Analysis of Results}\label{sect:validation:analysis}

\paragraph{Hypothesis H1}
The first analysis concerns hypothesis H1, whether the additions of RationalGRL -- arguments, critical questions and determination of the status of arguments -- are a useful addition to standard goal modeling in GRL. Table~\ref{table:survey:table2} summarizes the average scores, standard deviations, p-values for a right-tailed t-test, effect size and statistical power for the relevant questions Q5-Q7. As indicated by the p-values, which are all lower than 0.05, all the averages are significantly higher than the median of 3 with a significance level ($\alpha$) of 0.05, which means that we can accept H1$_{a}$ and reject H1$_{0}$ for all the separate questions Q5-Q7. In other words, the fact that the participants rate the additions of RationalGRL as, on average, (very) useful is very unlikely to be due to chance.

@@

\paragraph{Hypothesis H3}
The third hypothesis H3 is about the usability of the RationalGRL tool. It is different from the other two hypotheses in that we do not expect or claim that the usability of the tool is very good (or bad). Rather, because in the RationalGRL framework, the tool is more a means than an end in itself, we hypothesize that the too, is neither easy or difficult to use. Table~\ref{table:survey:table1} summarizes the average scores, standard deviations and p-values for a two-tailed t-test for questions Q1-Q4. As indicated by the p-values which are all higher than $\alpha=0.05$, none of the results is statistically significant, that is, the fact that there is a difference between the median of 3 and the average scores for the questions could very well be due to chance. This means we cannot reject the null hypothesis H3$_{0}$ and that, hence, the usability of the RationalGRL tool is neither (very) easy or (very) difficult. This corresponds to the answers to the open questions the participants gave: respondents were positive about the fact that it was very easy to get started with the tool and most comments on the tool were the type of comments that are expected from a prototype implementation.

\subsection{Threats to validity}
Internal validity is about the validity of the experiment results given our experiment setup and interpretation of findings. The main threat in this regard is that the modeling task the participants were asked to do is not realistic. The task is quite small, and based on an existing transcript of a requirements discussion. In a realistic setting, users of RationalGRL would model much more complex requirements together with the stakeholders. That said, the discussions analyzed in our case study (Section \ref{sect:gmas}) were of a realistic size and included many arguments and critical questions, so these concepts seem to take a central role in goal modeling and requirements engineering.

Another threat to internal validity is that many of the participants had no experience with goal modeling languages at all. Hence, the perceived usefulness of the arguments could be influenced by the fact that the participants, for the first time, were given a goal modeling language to perform structured requirements analysis. In order to test the usefulness of arguments and critical questions alone, one would have to compare a test and control group, where the test group uses RationalGRL and the control group uses normal GRL.

@@

\begin{enumerate}
\item \emph{Arguments and automatically determining acceptability are useful.} Participants were enthusiastic about adding arguments to a goal modeling language with a clear formal underpinning, and they believed the argumentation semantics we use in the tool are intuitive. This is a positive signal for the addition of arguments to goal modeling languages, and also for our formal approach, as just adding arguments as, for example, labels on goals and tasks does not allow one to compute the status of elements in a goal model in the way our approach does. Furthermore, automatically determining the status of elements will perhaps have a bigger impact when working with larger, more complex goal models, for which manually determining the impact of arguments is not feasible.
\item \emph{Critical questions need to be clear and have a prominent place in the process of building a goal model.} While participants overall found the critical questions (CQ's) useful, their impact was markedly less than the addition of arguments. This can have various causes. Participants mentioned they did not use the details pane of the tool, which can be used to ask and answer critical questions, that much. Furthermore, it was also mentioned that the difference between different questions was not always clear. Finally, the experiment task, modeling an existing discussion, may also have influenced the CQ's perceived usefulness. The CQ's are meant to be asked during the actual RationalGRL development process in which goal models are constructed and critically analyzed from scratch. 
\item \emph{High cognitive overhead.} A concern that was raised often in our empirical evaluation is that, in its current form, RationalGRL has a relatively high cognitive overhead. Goal modeling is by itself already a cognitively high-effort activity, and the fact that we add more elements to the language does not improve this. 
\end{enumerate}
